Language Skills in an Ethnically Segmented Labour market: Estonia 1989 -- 2012

### 2018-07-20 paper/ijm.tex
better wage residuals table and discussion in the 7th section

\thanksfootextra{}{\hfill}
\setlength{\thanksmargin}{1em}
\setlength{\thanksmarkwidth}{0.75em}

@@

When the central Soviet authority weakened during the late 1980, the
Estonian-speaking community was quick to organize and establish a
nation-state in 1991 where
Estonian was granted status as the sole official language.  The
above-mentioned ``one-sided bilingualism'' slowly started turning
around.  The universities quickly moved to Estonian-only curriculum,
Estonian language become a large-scale compulsory subject in Russian
schools while teaching Russian in Estonian-language schools declined.
Since 2007, all Russian high-schools have to teach part of the
curriculum in Estonian \citep{lindemann+saar2011Russian2ndGeneration}.

@@

The most surprising result in our analysis so far--the fact that
Russian men earn virtually no Estonian language premium while women do--warrants a
deeper look at the data.  We compute the average wage for the Russian male wage-earners
along the three
dimensions: Estonian language skills, residence in Tallinn metro area,
and education.  The metropolitan area is one of the two major regions
where the Russian-speaking population is concentrated, and it's
booming economy may be very different from that in the North-East.
Even more, while Russians form a large minority in and around Tallinn,
they are in majority in the North-East.  
In addition, to correct very strong wage and language skill trends through these
years, we also present the average
residuals from an OLS regression that explains log wage by year fixed
effects (Table~\ref{tab:loc-lang-edu}). 
\fxwarning{ot: Sven, can you redo this table?  Use exactly the same
  sample as you used.  Would also be good to put both et1 and et0 on
  the same row}
\fxwarning{SKB: Will try. Although I still do not understand, what the residual is telling me and why the wages are so high. Are the wages in EEK? What does ET stand for? The table needs more explanation or at least a proper legend.}

The table shows that the picture is in fact quite complex.  We can see
that there is little to no gain from Estonian skills for the
largest group, high-school educated workers.  Even more, outside of
the metropolitan area, the lowest and highest education groups
have in fact negative language premium.  However, these two groups
enjoy quite a noticeable premium in the Tallinn region.  The
opposite-signed effects in for these two education groups suggests
that a
selective internal migration may play a role.

% latex table generated in R 3.4.4 by xtable 1.8-2 package
% Fri Jul  6 15:59:47 2018
\begin{table}[ht]
  \centering
  \caption{Average wage and time-trend corrected residual across
    location and education groups for Russian men}
  \begin{tabular}{lrlrrr}
    \toprule
    Tallinn & education & ET & wage & residual & \# obs \\ 
    \midrule
    Tallinn & $<$ HS & 0 & 4313.38 & -0.19 & 255 \\ 
    Tallinn & $<$ HS & 1 & 5351.11 & -0.05 & 219 \\ 
    Tallinn & HS     & 0 & 5803.21 & 0.05 & 1292 \\ 
    Tallinn & HS      & 1 & 5666.77 & 0.05 & 1296 \\ 
    Tallinn & college & 0 & 6562.57 & 0.16 & 212 \\ 
    Tallinn & college & 1 & 8453.14 & 0.30 & 518 \\ 
    other   & $<$ HS & 0 & 3877.84 & -0.13 & 461 \\ 
    other   & $<$ HS & 1 & 4496.62 & -0.18 & 230 \\ 
    other   & HS     & 0 & 4838.65 & -0.09 & 2233 \\ 
    other   & HS      & 1 & 5351.51 & -0.05 & 960 \\ 
    other   & college & 0 & 6283.38 & 0.17 & 278 \\ 
    other   & college & 1 & 6072.06 & 0.13 & 263 \\ 
    \bottomrule
  \end{tabular}
  \label{tab:loc-lang-edu}
\end{table}
\fxwarning{SKB: Copied and modified table above. }
\begin{table}[ht]
	\centering
	\caption{Average wage and time-trend corrected language parameter across
		location and education groups for Russian men}
	\begin{tabular}{lrlrrr}
		\toprule
		Tallinn & education & ET & wage & parameter & \# obs \\ 
		\midrule
		Tallinn & $<$ HS & 0 & 334 & -0.16\onepc & 255 \\ 
		Tallinn & $<$ HS & 1 & 424 & -0.08 & 219 \\ 
%		Tallinn & HS     & 0 &476 & 0.05 & 1292 \\ 
		Tallinn & HS      & 1 & 512 & -0.05 \fivepc & 1296 \\ 
		Tallinn & college & 0 & 520 & -0.02 & 212 \\ 
		Tallinn & college & 1 & 737 & 0.05 & 518 \\ 
		Ida-Viru & $<$ HS & 0 & 396 &  -0.20\onepc & 379 \\ 
		Ida-Viru & $<$ HS & 1 & 269 & -0.23\onepc & 20 \\ 
		Ida-Viru & HS     & 0 & 337 & -0.23\onepc & 1889 \\ 
		Ida-Viru & HS      & 1 & 368 & -0.21\onepc & 252 \\ 
		Ida-Viru & college & 0 & 451 & -0.07 & 246 \\ 
		Ida-Viru & college & 1 & 493 & -0.06 & 154 \\ 
		other   & $<$ HS & 0 & 279 & -0.23\onepc & 83 \\ 
		other   & $<$ HS & 1 & 334 & -0.19\onepc & 212 \\ 
		other   & HS     & 0 & 378& -0.13\onepc & 353 \\ 
		other   & HS      & 1 & 389 & -0.09\onepc & 713\\ 
		other   & college & 0 & 788 & 0.03 & 32\\ 
		other   & college & 1 & 414 & -0.11 & 110 \\ 
		\bottomrule
	\end{tabular}
	\label{tab:loc-lang-edu_m3}
\end{table}

\fxwarning{SKB: Added graph showing the same parameter estimates. Maybe a better formatted graph like this shows the story better than a table}
\begin{figure}[hbtp]
	\centering
	\caption{Coefficient plot of Model 3 for Russian Men separated by Region and Education}
	\label{fig:russianmenlanguagecoefsm3}
	\includegraphics[width=1.0\linewidth]{russian_men_language_coefs_m3}
\end{figure}

We can conclude the table~\ref{tab:loc-lang-edu} that our
most surprising finding, no positive effect of Estonian language
skills for Russian men, is primarily driven by workers with
high-school degree.  It also suggests that highly educated workers in
the metropolitan area enjoy, in fact, a substantial language premium.
However, this group is not very large.

The previous results can be explained by two hypotheses: first,
Russian--speaking men, at least those with high-school degree as their
highest completed education, are working
primarily in Russian language environment (ethnic segregation) while this is not true for
women; and second, Russian-speaking men are working in jobs with
little communication requirements (gender segregation).

To shed some additional light on these hypotheses, we use data
from 2008 Integration Monitor Survey.  We focus on
two questions: ``which is the language of communication at your
workplace?'' and ``which languages do you need at work''.  The first
question addresses the ethnic segregation while the second focuses on
the need of communication.

@@

Table~\ref{tab:environment_descriptive} shows the proportion of
Russian respondents who use Estonian as the main communication language at work, and
who need Estonian and English at work.  As we see, women are more likely to
use Estonian while the opposite is true for English.
Table~\ref{tab:environment_model} presents the regression results
where we estimate the likelihood of having the need and using Estonian at work
while
controlling for gender and other individual
characteristics, including education and geographic location.

@@

We document that all analyzed language skills are associated with
unemployment and wage, however in a different way. Our most
intriguing observation is that the largest group of
Russian men, high-school graduates, earn virtually no income premium if fluent in Estonian,
the majority and the official language of the country, while ability
to speak the language is related to a substantially
lower unemployment rate.
Our analysis supports \citet{YaoandOurs2015}, and \citet{Lindemann2013} findings that
the labour market is segmented between ethnic and gender lines.
However, we find that ethnic segregation is less prevalent in the male
labour market.
This suggests for majority of Russian speaking men, Estonian
language gives a distinct advantage by facilitating hiring by
Estonian-dominated employers.  However, our outcome also supports 
\citet{leppik+vihalemm2015JofBaltStud} conclusion that
these
jobs are associated with little career mobility.  There is also
anecdotal evidence suggesting that offices are
typically manned by ethnic Estonians while manufacturing units have a
large share of Russian workers.
One can speculate that this outcome is related to a substantial population share
with negative attitudes toward Russians \citep{korts2009JofBaltStud}.
On a more positive tone, we find
that this may not be the case for highly educated workers.

>>>>

\DeclareUnicodeCharacter{2010}{-}% support older LaTeX versions
\thanksfootextra{}{\hfill}
\setlength{\thanksmargin}{1em}
\setlength{\thanksmarkwidth}{0.75em}

@@

When the central Soviet authority weakened during the late 1980, the
Estonian-speaking community was quick to organize and establish a
nation-state in 1991 where
Estonian was granted status as the sole official language.  The
above-mentioned ``one-sided bilingualism'' slowly started turning
around.  The universities quickly moved to Estonian-only curriculum,
Estonian language become a large-scale compulsory subject in Russian
schools while teaching Russian in Estonian-language schools declined.
Since 2007, all Russian high schools have to teach part of the
curriculum in Estonian \citep{lindemann+saar2011Russian2ndGeneration}.

@@

The most surprising result in our analysis so far--the fact that
Russian men earn virtually no Estonian language premium while women do--warrants a
deeper look at the data.  We split the average wage for the Russian male wage-earners
along the three
dimensions: Estonian language skills (\emph{EST 0} and \emph{EST 1}),
residence in Tallinn metro area (\emph{Tallinn} and \emph{Other}),
and education (\emph{< HS}, \emph{HS}, \emph{College Degree}).  The metropolitan area is one of the two major regions
where the Russian-speaking population is concentrated, and it's
booming economy may be very different from the other such region, the
North-East (Ida-Virumaa), and the rest of the country.
Even more, while Russians form a large minority in and around Tallinn,
they are in majority in the North-East.  
In order to correct for the strong wage and language skill trends
through the observation period, we do not present the average wage,
but instead the respective OLS residuals where we explain the wage by year fixed
effects and include no other explanatory variables (Table~\ref{tab:loc-lang-edu}).

The table shows that the picture is in fact complex.  We can see
that there is little to no gain from Estonian skills for the
largest group of workers, those with a high school degree.  Even more, outside of
the metropolitan area, the lowest and highest education groups
have in fact negative language premium.  However, these two groups
enjoy quite a noticeable premium in the Tallinn region.  The
opposite-signed effects in for these two education groups suggests
that a
selective internal migration may play a role.

% latex table generated in R 3.4.4 by xtable 1.8-2 package
% Fri Jul  6 15:59:47 2018
\begin{table}[ht]
  \centering
  \caption{Average wage and time-trend corrected residual across
    location and education groups for Russian men}
  \begin{tabular}{llrrrr}
    \toprule
    & & &\multicolumn{2}{c}{Log Wage}\\
    Location & Education & \# Observations & EST 0 & EST 1 & Premium\\
    \midrule
    Tallinn & $<$ HS & 399 & -0.20 & -0.01 & 0.19 \\ 
    Tallinn & HS & 2164 & 0.07 & 0.04 & -0.03 \\ 
    Tallinn & College Degree & 610 & 0.15 & 0.30 & 0.15 \\ 
    Other & $<$ HS & 501 & -0.12 & -0.12 & -0.01 \\ 
    Other & HS & 2658 & -0.10 & -0.07 & 0.03 \\ 
    Other & College Degree & 429 & 0.17 & 0.11 & -0.06 \\ 
    \bottomrule
  \end{tabular}
  \begin{flushleft}
    Notes: \emph{Log Wage} refers to regression residual where log
    wage is explained by year fixed effects.  \emph{EST0} and
    \emph{EST 1} are mean values for those who cannot, and who can,
    speak Estonian; $\text{Premium} = \text{EST 1} - \text{EST 0}$.
    Education level \emph{< HS} refers to less than high school
    degree, \emph{HS} is the high school degree as the highest
    completed education level.
  \end{flushleft}
  \label{tab:loc-lang-edu}
\end{table}

We can conclude the table~\ref{tab:loc-lang-edu} that our
most surprising finding, no positive effect of Estonian language
skills for Russian men, is primarily driven by workers with
high school degree.  It also suggests that highly educated workers in
the metropolitan area enjoy, in fact, a substantial language premium.
However, this group is not very large.

The previous results can be explained by two hypotheses: first,
Russian--speaking men, at least those with high school degree as their
highest completed education level, are working
primarily in Russian language environment (ethnic segregation) while this is not true for
women; and second, Russian-speaking men are working in jobs with
little communication requirements (gender segregation).
To shed some additional light on these hypotheses, we use data
from 2008 Integration Monitor Survey.  We focus on
two questions: ``which is the language of communication at your
workplace?'' and ``which languages do you need at work''.  The first
question addresses the ethnic segregation while the second focuses on
the need of communication.

@@

Table~\ref{tab:environment_descriptive} shows the proportion of
Russian respondents who use Estonian as the main communication language at work, and
who need Estonian and English at work.  As we see, women are more likely to
use Estonian while the opposite is true for English.
Table~\ref{tab:environment_model} presents the corresponding regression results
where we estimate the likelihood of having the need, and using, Estonian at work
while
controlling for gender and other individual
characteristics, including education and geographic location.

@@

We document that all analyzed language skills are associated with
unemployment and wage, however in a different way. Our most
intriguing observation is that the largest group of
Russian men, high school graduates, earn virtually no income premium if fluent in Estonian,
the majority and the official language of the country, while ability
to speak the language is related to a substantially
lower unemployment rate.
Our analysis supports \citet{YaoandOurs2015}, and \citet{Lindemann2013} findings that
the labour market is segmented between ethnic and gender lines.
However, we find that ethnic segregation is less prevalent in the male
labour market.
This suggests for majority of Russian speaking men, Estonian
language gives a distinct advantage by facilitating hiring by
Estonian-dominated employers.  However, our outcome also supports 
\citet{leppik+vihalemm2015JofBaltStud} conclusion that
these
jobs are associated with little career mobility.  There is also
anecdotal evidence suggesting that offices are
typically manned by ethnic Estonians while manufacturing units have a
large share of Russian workers.
One can speculate that this outcome is related to a substantial population share
with negative attitudes toward Russians \citep{korts2009JofBaltStud}.
On a more positive tone, we find
that this may not be the case for highly educated workers.
